---
title: "Analyzing Trends in House of Representatives from 2016-2020"
subtitle: "INFO 526 - Project 1"
author: "The Power of The Voters"
format: 
  html:
    embed-resources: true
toc: true
execute:
  warning: false
  message: false
  results: hide 
  echo: false
editor: visual
---

# Abstract

This study explores US House election results from 1976 to 2022, focusing on US House of Representative trends from 2016-2022, and state-wide trends in Arizona from 2012-2022. From 2016-2020, the data was wrangled into three subsets, one for each cycle. To simplify the analysis, all parties that are outside of Republican and Democrat were grouped into a new generalized party called "Other". These subsets were then displayed on a US map, where the fill of the state indicates which party has the house majority, while the label of the state indicates which party had the popular vote majority. This revealed insights into how the electoral system can lead to politicians that may not represent the sentiment of the population of the state. Zooming in on the state of Arizona, voting trends were analyzed over a longer time frame of 10 years, from 2012 to 2022. The mode of voting as well as the change in party in each district was visualized to create insights into how mode of voting will influence the winning candidate, and how that influences the winning party in each district.

Insights into voting trends are very important metrics for politicians and the general population as it can be directly be related to other aspects of life such as the economy and global politics. These insights can then help politicians target states where voter sentiment can make or break a politicians campaign.

The analysis focused on time-series analysis as the variables were view over time and the changes noted as valuable insights. The limitations for this project include the assumption that the conglomeration of minor parties will create a larger third party that will vote different from Democrat and Republican, though this is not the case in real life, where some minor parties will align closer to the larger parties and some will votes on their own ideas. Ulitmately, This study provides insights into US voting patterns and the impact of election results on future voting trends.

```{r}

#setup
if(!require(pacman))
  install.packages("pacman") #utilize pacman to load each package, while ensuring nothing is redownloaded unnecesarily
options(timeout = 1200) #sf takes forever to download so we have to increase the timeout time

pacman::p_load(tidyverse, #Add new packages as necessary
               here,
               devtools,
               dlookr,
               dplyr,
               usmap,
               sf,
               ggrepel,
               scales,
               usmapdata, 
               grid,
               ggpubr
               )
remotes::install_github("ropensci/USAboundaries")
remotes::install_github("ropensci/USAboundariesData")

```

# Introduction

## Introducing the Dataset

## EDA

```{r}
elec <- readr::read_csv('data/house.csv')
describe(elec)
elec16to20 <- elec |>
  filter(between(year,2016,2020))  #filter years to between 2016 and 2020

#ggplot(elec) |> #create a plot to show change house elections from entire dataset


```

# Question 1: How did congressional voting trends change from 2016-2020

## Approach

```{r}
# Data Wrangling
 
for (date in c(2016, #aggregate state mode of party to determine which party has the 
               2018, #majority
               2020)) {
  result <- elec16to20 |>
    filter(year == date) |>
    group_by(state) |>
    summarize(winnerByHouseMajority = names(which.max(table(party))))
  assign(paste0('partyMaj', 
                date), 
         result)
}
  
for(date in c(2016,2018,2020)) { #for each state determine which party got the most votes
  result <- elec16to20 |>
    filter(year == date) |>
    group_by(state) |>
    summarize(winnerByTotalVotes = party[which.max(candidatevotes)], 
              partyVotes = sum(candidatevotes)) |>
    mutate(fips = fips(state)) #necessary for plotting on US map only on this plot because we will merge
  assign(paste0('genMaj',
                date), 
         result)
}

data2016 <- merge(genMaj2016,partyMaj2016) |>
  merge(centroid_labels(regions = c('states')), by = "fips") #adds in data for centroid of each state for labeling
data2018 <- merge(genMaj2018, partyMaj2018) |> 
  merge(centroid_labels(regions = c('states')), by = "fips")
data2020 <- merge(genMaj2020, partyMaj2020) |>
  merge(centroid_labels(regions = c('states')), by = "fips")



```

We analyze the election data from 2016, 2018 and 2020 to determine the party that has the majority in each state based on the house of representatives. We calculate the party with majority of seats in the house for each state in each election year. It shows that party that received the most total votes in each state for each year. Which tells us the most total votes in each state for each year. The analysis provides insights into the party distribution and voting trends across states in the specified election years.

## Analysis

```{r  fig.width = 14, fig.height = 34 }
#general plot just to get used to the function. We can add layers like a ggplot object
#For Question 1 we want three maps of the US where each state is colored based on the winning party, then also a number for each state corresponding to the party that had the most votes


#  us2016 <- plot_usmap(data = data2016, 
#            values = "winnerByHouseMajority", 
#            labels = FALSE) + 
#   scale_fill_manual(name = 'Winner by House Majority', 
#                     values = c("REPUBLICAN" = 'red' , "DEMOCRAT" = 'blue'))+ 
#   geom_label_repel(data = data2016,
#                    aes(label = abbr,
#                        geometry = geom,
#                        color = winnerByTotalVotes),
#                    stat = 'sf_coordinates',
#                    size = 5,
#                    force = 0.01, 
#                    min.segment.length = 0) +
#   scale_color_manual(name = 'Winner by Total Votes', values = c("REPUBLICAN"= "red", "DEMOCRAT"="blue")) +     guides(color = guide_legend(title = "Winner by Total Votes")) + 
#   labs(
#     title = 'House Majority vs Popular Vote Majority',
#     subtitle = "By State in 2016",
#     #caption = "Teal & Harmon | Tidy Tuesday 2023"
#   ) + 
#   theme(
#     plot.title = element_text(size = 25, hjust = 0.5, vjust = -10),
#     plot.subtitle = element_text(size = 24, hjust = 0.5, vjust = -10),
#     #plot.caption = element_text(size = 20, hjust = 0.5),
#     legend.text = element_text(size = 12, hjust = 0),
#     legend.title = element_text(size = 16, hjust= 0),
#     legend.position = c(0.6,0.03)
#   )
# 
# 
# us2018 <- plot_usmap(data = data2018, 
#            values = "winnerByHouseMajority", 
#            labels = FALSE) + 
#   scale_fill_manual(name = 'Winner by House Majority', 
#                     values = c("REPUBLICAN" = 'red' , "DEMOCRAT" = 'blue'))+ 
#   geom_label_repel(data = data2016,
#                    aes(label = abbr,
#                        geometry = geom,
#                        color = winnerByTotalVotes),
#                    stat = 'sf_coordinates',
#                    size = 5,
#                    force = 0.01, 
#                    min.segment.length = 0) +
#   scale_color_manual(name = 'Winner by Total Votes', values = c("REPUBLICAN"= "red", "DEMOCRAT"="blue")) +     guides(color = guide_legend(title = "Winner by Total Votes")) + 
#   labs(
#     title = 'House Majority vs Popular Vote Majority',
#     subtitle = "By State in 2018",
#     #caption = "Teal & Harmon | Tidy Tuesday 2023"
#   ) + 
#   theme(
#     plot.title = element_text(size = 25, hjust = 0.5, vjust = -10),
#     plot.subtitle = element_text(size = 24, hjust = 0.5, vjust = -10),
#     #plot.caption = element_text(size = 20, hjust = 0.5),
#     legend.text = element_text(size = 12, hjust = 0),
#     legend.title = element_text(size = 16, hjust= 0),
#     legend.position = c(0.6,0.03)
#   )
# 
# us2020 <- plot_usmap(data = data2020, 
#            values = "winnerByHouseMajority", 
#            labels = FALSE) + 
#   scale_fill_manual(name = 'Winner by House Majority', 
#                     values = c("REPUBLICAN" = 'red' , "DEMOCRAT" = 'blue'))+ 
#   geom_label_repel(data = data2016,
#                    aes(label = abbr,
#                        geometry = geom,
#                        color = winnerByTotalVotes),
#                    stat = 'sf_coordinates',
#                    size = 5,
#                    force = 0.01, 
#                    min.segment.length = 0,
#                    force_pull = 0.5) +
#   scale_color_manual(name = 'Winner by Total Votes', values = c("REPUBLICAN"= "red", "DEMOCRAT"="blue")) +     guides(color = guide_legend(title = "Winner by Total Votes")) + 
#   labs(
#     title = 'House Majority vs Popular Vote Majority',
#     subtitle = "By State in 2020",
#     caption = "Teal & Harmon | Tidy Tuesday 2023"
#   ) + 
#   theme(
#     plot.title = element_text(size = 25, hjust = 0.5, vjust = -10),
#     plot.subtitle = element_text(size = 24, hjust = 0.5, vjust = -10),
#     plot.caption = element_text(size = 20, hjust = 0.5),
#     legend.text = element_text(size = 12, hjust = 0),
#     legend.title = element_text(size = 16, hjust= 0),
#     legend.position = c(0.6,0.03)
#   )
# 
# ggarrange(us2016, us2018, us2020, ncol = 1)

```

### 2016

```{r  fig.width = 14, fig.height = 34}

#| label: Data Visualization

#general plot just to get used to the function. We can add layers like a ggplot object
#For Question 1 we want three maps of the US where each state is colored based on the winning party, then also a number for each state corresponding to the party that had the most votes


 us2016 <- plot_usmap(data = data2016, 
           values = "winnerByHouseMajority", 
           labels = FALSE) + 
  scale_fill_manual(name = 'Winner by House Majority', 
                    values = c("REPUBLICAN" = 'red' , "DEMOCRAT" = 'blue'))+ 
  geom_label_repel(data = data2016,
                   aes(label = abbr,
                       geometry = geom,
                       color = winnerByTotalVotes),
                   stat = 'sf_coordinates',
                   size = 5,
                   force = 0.01, 
                   min.segment.length = 0) +
  scale_color_manual(name = 'Winner by Total Votes', values = c("REPUBLICAN"= "red", "DEMOCRAT"="blue")) +     guides(color = guide_legend(title = "Winner by Total Votes")) + 
  labs(
    title = 'House Majority vs Popular Vote Majority',
    subtitle = "By State in 2016",
    caption = "Teal & Harmon | Tidy Tuesday 2023"
  ) + 
  theme(
    plot.title = element_text(size = 25, hjust = 0.5, vjust = -10),
    plot.subtitle = element_text(size = 24, hjust = 0.5, vjust = -10),
    plot.caption = element_text(size = 20, hjust = 0.5),
    legend.text = element_text(size = 12, hjust = 0),
    legend.title = element_text(size = 16, hjust= 0),
    legend.position = c(0.6,0.03)
  )

us2016
```

We generate a map of united states for the year 2016, Where each state is colored based on the majority party in the House of Representatives .We also included labels for each state indicating the party that received the most votes. The map uses different colors to represent the Republican Party and Democrat Party. We can see out of 50 states in US Democrat Party has won 40 states and the rest are won by Republican Party in the year 2016.

### 2018

=======

```{r fig.width = 14, fig.height = 14 }

us2018 <- plot_usmap(data = data2018,
           values = "winnerByHouseMajority",
           labels = FALSE) +
  scale_fill_manual(name = 'Winner by House Majority',
                    values = c("REPUBLICAN" = 'red' , "DEMOCRAT" = 'blue'))+
  geom_label_repel(data = data2016,
                   aes(label = abbr,
                       geometry = geom,
                       color = winnerByTotalVotes),
                   stat = 'sf_coordinates',
                   size = 5,
                   force = 0.01,
                   min.segment.length = 0) +
  scale_color_manual(name = 'Winner by Total Votes', values = c("REPUBLICAN"= "red", "DEMOCRAT"="blue")) + guides(color = guide_legend(title = "Winner by Total Votes")) +
  labs(
    title = 'House Majority vs Popular Vote Majority',
    subtitle = "By State in 2018",
    caption = "Teal & Harmon | Tidy Tuesday 2023"
  ) +
  theme(
    plot.title = element_text(size = 25, hjust = 0.5, vjust = -10),
    plot.subtitle = element_text(size = 24, hjust = 0.5, vjust = -10),
    plot.caption = element_text(size = 20, hjust = 0.5),
    legend.text = element_text(size = 12, hjust = 0),
    legend.title = element_text(size = 16, hjust= 0),
    legend.position = c(0.6,0.03)
  )

us2018
```

We generate a map of united states for the year 2018, Where each state is colored based on the winning party in the House of Representatives. We also included labels for each state indicating the party that received the most votes. The map uses different colors to represent the Republican Party and Democrat Party. We can see out of 50 states in US Democrat Party has won 42 states and Republican Party has won only in 1 state that is NC in 2018.

### 2020

```{r fig.width = 14, fig.height = 14 }


us2020 <- plot_usmap(data = data2020,
           values = "winnerByHouseMajority",
           labels = FALSE) +
  scale_fill_manual(name = 'Winner by House Majority',
                    values = c("REPUBLICAN" = 'red' , "DEMOCRAT" = 'blue'))+
  geom_label_repel(data = data2016,
                   aes(label = abbr,
                       geometry = geom,
                       color = winnerByTotalVotes),
                   stat = 'sf_coordinates',
                   size = 5,
                   force = 0.01,
                   min.segment.length = 0,
                   force_pull = 0.5) +
  scale_color_manual(name = 'Winner by Total Votes', values = c("REPUBLICAN"= "red", "DEMOCRAT"="blue")) +     guides(color = guide_legend(title = "Winner by Total Votes")) +
  labs(
    title = 'House Majority vs Popular Vote Majority',
    subtitle = "By State in 2020",
    caption = "Teal & Harmon | Tidy Tuesday 2023"
  ) +
  theme(
    plot.title = element_text(size = 25, hjust = 0.5, vjust = -10),
    plot.subtitle = element_text(size = 24, hjust = 0.5, vjust = -10),
    plot.caption = element_text(size = 20, hjust = 0.5),
    legend.text = element_text(size = 12, hjust = 0),
    legend.title = element_text(size = 16, hjust= 0),
    legend.position = c(0.6,0.03)
  )

us2020
```

We generate a map of united states for the year 2020, Where each state is colored based on the wining party in the House of Representatives. We also included labels for each state indicating the party that received the most votes. The map uses different colors to represent the Republican Party and Democrat Party. We can see out of 50 states in US Democrat Party has lost 13 states which is the most states by comparing 2016, 2018 and 2020 election results. Also, Republican party has gained most states according to the data of previous years.

## Discussion

The visualization of the House of Representatives broken down by district majority and popular vote majority provides multiple insights. 1) The sentiment of the general population is often overlooked by the electoral (district) system. 2) During a single presidency, the sentiment of the people will change significantly 3) Well known trends can be seen such as midterm elections supporting the opposite party of the incumbent president.

# Question 2 : How often did change occur in House representation from the years 2012-2022 in the state of Arizona and which voting methods played a significant role in these elections?

## Approach

```{r}
# Data wrangling for district maps

#| label: Question 2 Data Wrangling

elec2 <- readr::read_csv('data/house_population_merged.csv')

elec12to22 <- elec2 |>
  filter(between(year,2012,2022))

ariz <- elec12to22 |>
  filter(state == "ARIZONA",
         !is.na(party))

for(date in c(2012, 2016, 2022)){
  result2 <- ariz |>
    filter(year == date) |>
    group_by(district) |>
    summarise(winParty = party[which.max(candidatevotes)])
  assign(paste0('distWin', date), result2)
}

```

The analysis will look at election results from 2012, 2016, and 2022 to see how Arizona's congressional district alignments have changed over time. The goal is to identify patterns of political representation inside the state's districts, detecting movements in party control over both short-term and decade-long periods. Furthermore, the analysis will look into the influence of various voting procedures, specifically their impact on election outcomes in Arizona.

## Analysis

### 2012

```{r}
#| label: 2012 District Map

# Get dataframe to map AZ state from USAboundaries
cd_az12 <- USAboundaries::us_congressional(resolution = "high", states = c("Arizona")) |> 
  st_sf() |>
  rename(district = "cd116fp")

# Change district variable type to numeric and rename it 'district'
cd_az12$district <- as.numeric(cd_az12$district)

# Join USAboundaries data with our dataset to determine party winners of each district
azMap12 <- full_join(cd_az12, distWin2012, by = 'district')

# Plot district map
azMap12 |>
  ggplot() +
  geom_sf(aes(fill = winParty)) +
  scale_fill_manual(name = "Party", values = c("blue", "red")) +
  labs(title = "Arizona Congressional Districts 2012") +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.75),
        panel.grid = element_blank(),
        panel.background = element_blank())

```

The resultant visualisation is a color-coded map of Arizona's congressional districts in 2012, based on the political party that won each seat. Each district is assigned a colour that represents the winning party: blue for Democrats and red for Republicans. The representation offers a visual picture of Arizona's political distribution in 2012, highlighting regions of Democratic and Republican strength and providing insights into regional political processes. This visualisation reveals the political leaning of most of the congressional districts in Arizona during that election cycle was towards the Democratic party.

### 2016

```{r}
#| label: 2016 District Map

# Get dataframe to map AZ state from USAboundaries
cd_az16 <- USAboundaries::us_congressional(resolution = "high", states = c("Arizona")) |> 
  st_sf() |>
  rename(district = "cd116fp")

# Change district variable type to numeric and rename it 'district'
cd_az16$district <- as.numeric(cd_az16$district)

# Join USAboundaries data with our dataset to determine party winners of each district
azMap16 <- full_join(cd_az16, distWin2016, by = 'district')

azMap16$change <- c("yes", "no", "no", "no", "no", "no", "no", "no", "no")

# Plot district map
azMap16 |>
  ggplot() + 
  geom_sf(aes(fill = winParty, alpha = change)) +
  scale_fill_manual(name = "Party", values = c("blue", "red")) +
  scale_alpha_discrete(range = c(0.5, 1)) +
  labs(title = "Arizona Congressional Districts 2016") +
  guides(alpha = FALSE) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.75),
        panel.grid = element_blank(),
        panel.background = element_blank())


```

The resultant visualisation is a color-coded map of Arizona's congressional districts in 2016, based on the political party that won each seat. Each district is assigned a colour that represents the winning party: blue for Democrats and red for Republicans. One key feature of this visualisation is the use of transparency to highlight shifts in political representations. Districts with a party change are displayed with a lower opacity, distinguishing them from those with no change, which keep full-colour saturation. This visualisation reveals that the political leaning of Cochise district during the 2016 election cycle has changed from Democratic party to Republican party.

### 2022

```{r}
#| label: 2022 District Map

# Get dataframe to map AZ state from USAboundaries
cd_az22 <- USAboundaries::us_congressional(resolution = "high", states = c("Arizona")) |> 
  st_sf() |>
  rename(district = "cd116fp")

# Change district variable type to numeric and rename it 'district'
cd_az22$district <- as.numeric(cd_az22$district)

# Join USAboundaries data with our dataset to determine party winners of each district
azMap22 <- full_join(cd_az22, distWin2022, by = 'district')

azMap22$change <- c("no", "no", "yes", "yes", "no", "yes", "no", "no", "no")

# Plot district map
azMap22 |>
  ggplot() + 
  geom_sf(aes(fill = winParty, alpha = change)) +
  scale_fill_manual(name = "Party", values = c("blue", "red")) +
  scale_alpha_discrete(range = c(0.5, 1)) +
  labs(title = "Arizona Congressional Districts 2022") +
  guides(alpha = FALSE) +
  theme(axis.text = element_blank(),
        axis.ticks = element_blank(),
        plot.title = element_text(hjust = 0.75),
        panel.grid = element_blank(),
        panel.background = element_blank())




```

## Discussion

From the Arizona congressional district maps we can see that there are not many changes that occur in the span of only 2-4 years, however there are significant changes that occur when comparing the district maps that are a decade apart. From the rest of the dataset we are able to see that this is often the case, it is not common for many districts to change the parties that represent them in a span of only one or two elections, however this doesn't necessarily mean that change is a rarity as we can see the significant change in district representation over the course of a decade.

# ANY CODE BELOW THIS SECTION ADD TO THE ABOVE SECTIONS

## EDA for voting methods

```{r}
#| label: Load new features into dataset

library(readxl)
house_with_New_features <- read_csv('data/house_with_New_features.csv', show_col_types = FALSE)
```

## Data wrangling for voting methods

```{r}
#| label: Data Wrangling Q.2

# Load necessary libraries
library(readxl)
library(dplyr)
library(ggplot2)

# Filter data for Arizona state, relevant columns, and years 2012-2022
voting_data_arizona <- house_with_New_features %>%
  filter(state == "ARIZONA", year >= 2012 & year <= 2022) %>%
  select(year, Type_of_Voting, RESULT)

# Count the number of candidates for each combination of Year, Type_of_Voting, and RESULT
voting_result_counts_arizona <- voting_data_arizona %>%
  group_by(year, Type_of_Voting, RESULT) %>%
  summarise(Count = n())

# Calculate the percentage of candidates for each combination
voting_result_counts_arizona <- voting_result_counts_arizona %>%
  group_by(year, Type_of_Voting) %>%
  mutate(Percentage = (Count / sum(Count)) * 100) %>%
  ungroup() %>%
  arrange(year, desc(Count))

# Custom color palette
custom_colors <- c("#009688", "#FF9800")

```

## Data Visualization

```{r}
#| label: Data Visualization Q.2

# Visualization - Enhanced Horizontally Stacked Bar Plot for Arizona with Percentage Labels

ggplot(voting_result_counts_arizona, aes(x = reorder(factor(Type_of_Voting, levels = unique(Type_of_Voting)), Count), y = Count, fill = RESULT)) +
  geom_bar(stat = "identity", color = "white") +
  geom_text(aes(label = paste0(round(Percentage, 1), "%")), 
            position = position_stack(vjust = 0.5),
            color = "black", size = 5, fontface = "bold") +  # Add percentage labels
  facet_wrap(~ year, scales = "free") +
  labs(title = "Impact of Types of Voting on House Representation in Arizona (2012-2022)",
       y = "Number of Candidates",
       x = "Type of Voting") +
  scale_fill_manual(values = custom_colors) +  # Set custom colors
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5,size = 14),  # Rotate labels by 90 degrees
        axis.text.y = element_text(size = 14),
        plot.title = element_text(hjust = 0.5, size = 18, face = "bold"),
        legend.title = element_text(size = 16),
        legend.text = element_text(size = 16),
        legend.position = "bottom",
        panel.grid.major.y = element_blank(),  # Remove horizontal grid lines
        panel.grid.minor.y = element_blank(),  # Remove horizontal grid lines
        panel.grid.major.x = element_blank(),  # Remove horizontal grid lines
        panel.grid.minor.x = element_blank(),  # Remove horizontal grid lines
        panel.background = element_rect(fill = "white"),  # Set panel background color
        plot.background = element_rect(fill = "#F5F5F5"),  # Set plot background color
        strip.background = element_rect(fill = "#E0E0E0", color = "white"),  # Set facet strip background color
        strip.text = element_text(face = "bold",size = 16),  # Set strip text size
        axis.title.y = element_text(size = 18),  # Set y-axis label size
        axis.title.x = element_text(size = 18)) +  # Set x-axis label size
  coord_flip()  # Flip the coordinates to create a horizontally stacked bar chart

```
