---
title: "Proposal title"
subtitle: "Proposal"
format: html
editor: visual
---

```{r}
#| label: load-pkgs
#| message: false

if(!require(pacman))
  install.packages("pacman")

pacman::p_load(tidyverse,
               here,
               dlookr)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false

tuesdata <- tidytuesdayR::tt_load('2023-11-07')
house <- tuesdata$house
```
The dataset, sourced from MIT Election Data and Science Lab (MEDSL), offers a comprehensive overview of US House elections.

This dataset contains observations for elections held over 47 years from 1976 to 2022, encompassing a total of 32,452 recorded events. Each event is represented as a row with 20 attributes as columns. These columns provide details including the year, state, district, political party, candidate's name, votes received, and various indicators such as whether it was a runoff election or if the candidate ran as a write-in.

Make sure to load the data and use inline code for some of this information.

## Questions

Question 1: Which party wins in each state for the general election on a map of the USA?

Question 2: How is the mode of voting influenced by district population and how does that influence who loses?

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
