---
title: "The Power of Voters" 
subtitle: "Proposal"
format: 
  html: 
   embed-resources: true
toc: true
editor: visual
execute:
  echo: false
  warning: false
---

```{r}
#| label: load-pkgs
#| message: FALSE
#| warning: false

if(!require(pacman))
  install.packages("pacman")

pacman::p_load(tidyverse,
               here,
               devtools,
               dlookr)

#devtools::install_github("choonghyunryu/dlookr")
```

## Dataset

#### US House Election Results

```{r}
#| label: load-dataset
#| message: false
#| warning: false

house <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-11-07/house.csv')

```

<<<< Updated upstream
| Variable         | Description                                      |
|------------------|--------------------------------------------------|
| `year`           | Year in which election was held                  |
| `state_po`       | State name                                       |
| `party`          | Party of the candidate                           |
| `mode`           | Mode of voting                                   |
| `candidatevotes` | No. of votes the candidate received.             |
| `fusion ticket`  | candidate will receive votes from multiple party |

```{r}
#| label: Summary Statistics
#| warning: false
#| message: false

# Summary statistics
# house |> dlookr::describe()
```

**About the dataset :**

The dataset,[US House Election Results](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-11-07) sourced from MIT Election Data and Science Lab (MEDSL), offers a comprehensive overview of US House elections.
=======
The dataset, sourced from MIT Election Data and Science Lab (MEDSL), offers a comprehensive overview of US House elections.
>>>>>>> Stashed changes

This dataset contains observations for elections held over 47 years from 1976 to 2022, encompassing a total of 32,452 recorded events. Each event is represented as a row with 20 attributes as columns. These columns provide details including the year, state, district, political party, candidate's name, votes received, and various indicators such as whether it was a runoff election or if the candidate ran as a write-in.

**Reason for choosing:**

The US House Election dataset offers a rich and comprehensive source of data, encompassing multiple variables such as election years, state information, candidate details, and voting outcomes. This abundance of data provides ample opportunities for rigorous analysis and exploration within the scope of the academic project

## Questions

Question 1: How do election outcomes vary by political party across different states in the USA over a 4-year period (1976-2022), and what demographic factors may influence these variations?

Question 2: How is the type of voting influenced by population and how does that influence who looses?

## Analysis plan

**For Question 1:**

PART 1:   How do election outcomes vary by political party across different states in the USA over a 4-year period (1976-2022)?                                                                    

*Variables Involved:*

-   **`state_po`**: The two-letter postal abbreviation for the state.

-   **`party`**: The political party of the candidate.

-   **`year`**: Year of the election, focusing on the most recent general election or a specific year of interest.

-   **`candidatevotes`**: The number of votes the candidate received.

Analysis Approach:

1.  **Filter Data**: Data selection from the general elections (**`stage`** variable) for the most recent year or a specific year of interest.

2.  **Determine Winners**: For each state, identify the party with the highest aggregate **`candidatevotes`** in the general election.

3.  **Visualization**: Create a map of the USA, coloring each state based on the winning party. This visualization will clearly depict the political landscape across the states for the selected election year.

PART 2: what demographic factors may influence these variations?

*Variables Involved:*
       
        Internal Data Source:
-   **`State_po`**: Name or code representing a U.S. state.
-   **`Party`**: Political affiliation of a candidate.
-   **`CandidateVotes`**: Number of votes received by a candidate.
-   **`Year`**: Year in which the election was held.
       
        Demographic Data from External Sources
-   **`Population Demographics`**: Characteristics of a population, such as size and density.
-   **`Voter Demographics`**: Characteristics of voters, including registration and turnout rates.
-   **`Income Levels`**: Distribution of income among residents.
-   **`Education Levels`**: Distribution of educational attainment.
-   **`Race/Ethnicity Distributions`**: Distribution of racial and ethnic groups within the population

Analysis Approach:

Scatter Plot for Demographic Analysis:

1)  **Filter Data**:
Combining election outcome data with demographic data, ensuring alignment based on common identifiers such as state names or codes. 

2)  **Visualization**
Generating a scatter plot, .with each data point representing a state and its election outcome correlated with the corresponding demographic factor like population, income, voter, education, race/ethnicity distributions

Heatmap for Correlation Analysis:

1)  **Filter Data**:
Combining election outcome data with demographic data, ensuring alignment based on common identifiers such as state names or codes. 

2)  **Visualization**:
Calculating  the correlation coefficients between the election outcomes (DV) and each demographic factor like population, income, voter, education, race/ethnicity distributions 
analyzing the heatmap, we can identify which demog.raphic factors have the strongest correlation with election outcomes and understand how they influence political dynamics.


***Key Points***:
•	The analysis integrates both internal and external data sources, combining election outcome data with demographic data obtained from external sources







**For Question 2:**

*Variables Involved:*

-   **`district`**: The congressional district number.

-   **`fusion ticket`** : allowing candidate to receive votes from multiple party.

-   **`runoff`** : second election held to determine a winner

-   **`special`** : Special election

-   **`party`**: The political party of the candidate.

-   **`candidatevotes`**: The number of votes the candidate received.

-   **`candidate`**: Name of candidate

*Variables to be Created:*

-   **`population`**: This will require merging external data containing population

***Analysis Approach:***

1.  **Merge External Data**: Incorporate population data from a reliable source (e.g. U.S. Census Bureau).

2.  **Analyze Voting Type by Population**: Examine the relationship between population and the type of voting, using statistical analysis .

3.  **Correlate with Election Outcomes**: Assess how the type of voting and population correlate with the election outcomes, focusing on which party tends to loose with certain characteristics.

***Visualization and Reporting:***

-   Use scatter plots or heat maps to show the relationship between population and mode of voting.

-   Create comparative visualizations to illustrate how these factors influence election outcomes, potentially using bar charts and pie chart
