---
title: "Proposal title"
subtitle: "Proposal"
format: html
editor: visual
echo : false 
warning : FALSE  
message : FALSE  
---

```{r}
#| label: load-pkgs
#| message: FALSE

if(!require(pacman))
  install.packages("pacman")

pacman::p_load(tidyverse,
               here,
               devtools,
               dlook)

# devtools::install_github("choonghyunryu/dlookr")
```

## Dataset

```{r}
#| label: load-dataset
#| message: FALSE
#| warrning: False


house <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-11-07/house.csv')


# Summary statistics
sumStats <- house |> dlookr::describe()
View(sumStats)

```

About the dataset: The dataset,[US House Election Results](https://github.com/rfordatascience/tidytuesday/tree/master/data/2023/2023-11-07) sourced from MIT Election Data and Science Lab (MEDSL), offers a comprehensive overview of US House elections.

This dataset contains observations for elections held over 47 years from 1976 to 2022, encompassing a total of 32,452 recorded events. Each event is represented as a row with 20 attributes as columns. These columns provide details including the year, state, district, political party, candidate's name, votes received, and various indicators such as whether it was a runoff election or if the candidate ran as a write-in.

Reason for choosing: The US House Election dataset offers a rich and comprehensive source of data, encompassing multiple variables such as election years, state information, candidate details, and voting outcomes. This abundance of data provides ample opportunities for rigorous analysis and exploration within the scope of the academic project

## Questions

Question 1: Which party wins in each state for the general election on a map of the USA?

Question 2: How is the mode of voting influenced by district population and how does that influence who loses?

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).
